---
title: "RMS Titanic"
author: "(Your name here)"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|------------------|-----------------------------|-------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look

<!-- -------------------------------------------------- -->

### **q1** Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
glimpse(df_titanic)
```

**Observations**:

-   The variables are Class, Sex, Age, Survived, and n.

### **q2** Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

-   Write your observations here
    -   The total number of passengers from the real-life data is estimated to be 2224, while the computation result from the data set above is 2201.
-   Are there any differences?
    -   Yes - the computed result is 1.03% less than the real data, which is pretty accurate.
-   If yes, what might account for those differences?
    -   The computed result might be derived from a different sample size.

    -   The limitation of technology as well as the documentation of the event that was dated to more than a century ago might have contributed to the uncertainties of the data in general.

### **q3** Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`
df_titanic %>% filter(Survived == "Yes") %>%
  ggplot() +
  geom_col(mapping = aes(x = Class, y = n, fill = Sex) , position = "dodge")
```

**Observations**:

-   In the first class, the number of survived female passengers is about twice of that of male passengers.
-   In the second class, the number of survived female passengers outweighs male passengers by a few times.
-   The number of survived female passengers is about the same compared to that of male passengers in the third class.
-   Very few female survivors came from the crew class, while around 180 male passengers were from the crew class.

# Deeper Look

<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### **q4** Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
df_prop %>% 
  filter(Survived == "Yes") %>%
  ggplot() +
  geom_col(mapping = aes(x = Class, y = Prop, fill = Sex), position = "dodge")
```

**Observations**:

-   Women have a higher survival rate in all classes and at all ages compared to men.

-   The survival rate in the first two classes are much higher than that in the third and the crew classes.

-   Is there anything *fishy* going on in your plot?

    -   While the number of survived female passengers is much smaller than the number of survived male passengers in the crew class, the proportion of survived female passengers is much higher than that of survived male passengers.

    -   The proportions of both female and male survivors in the first and second classes are one.

    -   In general, this plot is quite different from the graph that plots the total counts of the same groups.

### **q5** Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
df_prop %>% filter(Survived == "Yes") %>%
  ggplot() +
  geom_col(mapping = aes(x = Class, y = Prop, fill = Sex) , position = "dodge") +
  facet_wrap(~ Age)
```

**Observations**:

-   Women have a higher survival rate in all classes and at all ages compared to men.

-   The children in the first and second classes have survival rate of one, which is much higher compared to the adults in the first two classes.

-   There is no children in the crew class.

-   The plots of the first, second and third classes in q4 line up perfectly with the corresponding plots of Child in q5, while the plot of the crew class in q4 lines up with the crew class of Adult in q5.

-   If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.

    -   q4 didn't take care of the Age grouping in df_prop, and therefore takes the proportion results of either the Adult group or the Child group by chance (in this case, the Child group appears first in the data set and therefore the result of the Child group fills up the visual of q4 in the first three classes).
    -   No children are employed to be a crew member, so only the result of the Adult group can be reflected on the graph under q5.

# Notes

<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
