---
title: "Antibiotics"
author: "(Your name here)"
date: 2020-
output:
  github_document
editor_options: 
  markdown: 
    wrap: 72
---

*Purpose*: Creating effective data visualizations is an *iterative*
process; very rarely will the first graph you make be the most
effective. The most effective thing you can do to be successful in this
iterative process is to *try multiple graphs* of the same data.

Furthermore, judging the effectiveness of a visual is completely
dependent on *the question you are trying to answer*. A visual that is
totally ineffective for one question may be perfect for answering a
different question.

In this challenge, you will practice *iterating* on data visualization,
and will anchor the *assessment* of your visuals using two different
questions.

*Note*: Please complete your initial visual design **alone**. Work on
both of your graphs alone, and save a version to your repo *before*
coming together with your team. This way you can all bring a diversity
of ideas to the table!

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics
define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|------------------|--------------------------|-----------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight**
before the day of the class discussion of the challenge. See the
[Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true)
for more information.

```{r setup}
library(tidyverse)
library(ggrepel)
library(ggplot2) 
library(treemapify)
library(ggplotify)
```

*Background*: The data[1] we study in this challenge report the
[*minimum inhibitory
concentration*](https://en.wikipedia.org/wiki/Minimum_inhibitory_concentration)
(MIC) of three drugs for different bacteria. The smaller the MIC for a
given drug and bacteria pair, the more practical the drug is for
treating that particular bacteria. An MIC value of *at most* 0.1 is
considered necessary for treating human patients.

These data report MIC values for three antibiotics---penicillin,
streptomycin, and neomycin---on 16 bacteria. Bacteria are categorized
into a genus based on a number of features, including their resistance
to antibiotics.

```{r load}
## NOTE: If you extracted all challenges to the same location,
## you shouldn't have to change this filename
filename <- "./data/antibiotics.csv"

## Load the data
df_antibiotics <- read_csv(filename)
df_antibiotics %>% knitr::kable()
```

# Visualization

<!-- -------------------------------------------------- -->

### **q1** Prototype 5 visuals

To start, construct **5 qualitatively different visualizations of the
data** `df_antibiotics`. These **cannot** be simple variations on the
same graph; for instance, if two of your visuals could be made identical
by calling `coord_flip()`, then these are *not* qualitatively different.

For all five of the visuals, you must show information on *all 16
bacteria*. For the first two visuals, you must *show all variables*.

*Hint 1*: Try working quickly on this part; come up with a bunch of
ideas, and don't fixate on any one idea for too long. You will have a
chance to refine later in this challenge.

*Hint 2*: The data `df_antibiotics` are in a *wide* format; it may be
helpful to `pivot_longer()` the data to make certain visuals easier to
construct.

#### Visual 1 (All variables)

In this visual you must show *all three* effectiveness values for *all
16 bacteria*. You must also show whether or not each bacterium is Gram
positive or negative.

```{r q1.1}
# WRITE YOUR CODE HERE
df_antibiotics %>% 
  ggplot(aes(x = bacteria)) + 
  geom_point(aes(y = streptomycin, color = "streptomycin", size = "streptomycin", alpha = gram)) +
  geom_point(aes(y = penicillin, color = "penicillin", size = "penicillin", alpha = gram)) +
  geom_point(aes(y = neomycin, color = "neomycin", size = "neomycin", alpha = gram)) +
  ylab("MIC") +
  xlab("Bacteria") +
  scale_y_log10() + 
  guides(x =  guide_axis(angle = 90)) +
  ggtitle("MICs of 16 types of bacteria")
```

#### Visual 2 (All variables)

In this visual you must show *all three* effectiveness values for *all
16 bacteria*. You must also show whether or not each bacterium is Gram
positive or negative.

Note that your visual must be *qualitatively different* from *all* of
your other visuals.

```{r q1.2}
# WRITE YOUR CODE HERE
df_antibiotics %>%
  group_by(gram) %>%
  pivot_longer(
    c('penicillin', 'streptomycin', 'neomycin'),
    names_to = "antibiotics", 
    values_to = "MIC"
  ) %>%
  ggplot() +
  geom_col(mapping = aes(
    x = bacteria,
    y = MIC,
    fill = antibiotics
  ), position = position_dodge(0.8)
  ) +
  facet_wrap(~ gram) + 
  ylab("MIC") +
  xlab(" ") +
  scale_y_log10() +
  guides(x =  guide_axis(angle = 90)) +
  ggtitle("MICs of penicillin")

```

#### Visual 3 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`,
`streptomycin`, `neomycin`, `gram`), but you must still show *all 16
bacteria*.

Note that your visual must be *qualitatively different* from *all* of
your other visuals.

```{r q1.3}
# WRITE YOUR CODE HERE
ggplot(df_antibiotics, aes(x = bacteria)) +
  geom_line(aes(y = streptomycin, group = 1, color = 'Streptomycin'), linetype="dotted") +
  geom_line(aes(y = penicillin, group = 1, color = 'Penicillin'), linetype="dotted") +
  geom_line(aes(y = neomycin, group = 1, color = 'Neomycin'), linetype="dotted") +
  scale_color_manual(
    name = 'Y series', 
    values = c('Streptomycin' = 'brown', 'Neomycin' = 'navy', 'Penicillin' = 'orange')) +
  ylab("MIC") +
  xlab("Streptomycin") +
  scale_y_log10() +
  guides(x =  guide_axis(angle = 90)) +
  ggtitle("Streptomycin MICs of 16 types of bacteria") +
  theme(legend.position="top")
```

#### Visual 4 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`,
`streptomycin`, `neomycin`, `gram`), but you must still show *all 16
bacteria*.

Note that your visual must be *qualitatively different* from *all* of
your other visuals.

```{r q1.4}
# WRITE YOUR CODE HERE
df_antibiotics %>%
  ggplot(aes(x = streptomycin, y = penicillin)) +
  geom_point(aes()) +

# df_q4 %>%
#   ggplot(aes(x = reorder(bacteria, streptomycin))) +
#   geom_point(aes(y = streptomycin, color = 'Streptomycin')) +
#   geom_point(aes(y = neomycin, color = 'Neomycin')) +
#   geom_point(aes(y = penicillin, color = 'Penicillin')) +
  
  # geom_point(
  #   aes(
  #     xend = reorder(bacteria, streptomycin),
  #     y = 1,
  #     yend = streptomycin,
  #     color = 'Streptomycin')) +
  
  # geom_point(
  #   aes(
  #     xend = reorder(bacteria, streptomycin),
  #     y = 1,
  #     yend = neomycin,
  #     color = 'Neomycin')) +

  # geom_point(
  #   aes(
  #     xend = reorder(bacteria, streptomycin),
  #     y = 1,
  #     yend = penicillin,
  #     color = 'Penicillin')) +
  geom_text_repel(
    aes(label = bacteria),
    size = 2
  ) +

  ylab("penicillin") +
  xlab("streptomycin") +
  scale_y_log10() +
  scale_x_log10() +
  guides(x =  guide_axis(angle = 90)) +
  ggtitle("Streptomycin MICs of 16 types of bacteria") +
  # scale_color_manual(
  #   name = 'Y series', 
  #   values = c('Neomycin' = 'navy', 'Penicillin' = 'orange', 'Streptomycin' = 'brown')) +
  theme(legend.position="top")
```

#### Visual 5 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`,
`streptomycin`, `neomycin`, `gram`), but you must still show *all 16
bacteria*.

Note that your visual must be *qualitatively different* from *all* of
your other visuals.

```{r q1.5}
# WRITE YOUR CODE HERE
p1 <-  ggplot() +
  geom_col(
    data = df_antibiotics, 
    aes(
      x = bacteria, 
      y = penicillin, 
      color = gram
    ), fill = "orange") +
  ylab("MIC") +
  xlab(" ") +
  scale_y_log10() +
  guides(x =  guide_axis(angle = 90)) +
  ggtitle("MICs of penicillin")

p2 <- ggplot() +
  geom_col(
    data = df_antibiotics, 
    aes(
      x = bacteria, 
      y = streptomycin, 
      color = gram
    ), fill = "brown") +
  ylab("MIC") +
  xlab(" ") +
  scale_y_log10() +
  guides(x =  guide_axis(angle = 90)) +
  ggtitle("MICs of streptomycin")
  
p3 <- ggplot() +
  geom_col(
    data = df_antibiotics, 
    aes(
      x = bacteria, 
      y = neomycin, 
      color = gram
    ), fill = "navy") +
  ylab("MIC") +
  xlab(" ") +
  scale_y_log10() +
  guides(x =  guide_axis(angle = 90)) +
  ggtitle("MICs of neomycin")

library(patchwork)
p1 + theme(legend.position="top", axis.title.y = element_blank()) + 
  p2 + theme(legend.position="top", axis.title.y = element_blank()) + 
  p3 + theme(legend.position="top", axis.title.y = element_blank())
```

### **q2** Assess your visuals

There are **two questions** below; use your five visuals to help answer
both Guiding Questions. Note that you must also identify which of your
five visuals were most helpful in answering the questions.

*Hint 1*: It's possible that *none* of your visuals is effective in
answering the questions below. You may need to revise one or more of
your visuals to answer the questions below!

*Hint 2*: It's **highly unlikely** that the same visual is the most
effective at helping answer both guiding questions. **Use this as an
opportunity to think about why this is.**

#### Guiding Question 1

> How do the three antibiotics vary in their effectiveness against
> bacteria of different genera and Gram stain?

*Observations*

What is your response to the question above?

\- penicillin, streptomycin, neomycin,

-   From the first graph of all the bacteria types that have negative
    gram, penicillin has the highest MIC values and is the least
    effective, while neomycin has the lost MIC values and is the most
    effective.
-   From the second graph of all the bacteria types that have positive
    gram, penicillin has the lowest MIC values most of the time while
    neomycin has the lowest MIC values sometimes (less than penicillin).
    Streptomycin has relatively higher values of MIC compared to the
    other two types of antibiotics, so we can conclude that it is the
    least effective.
-   Generally speaking, the MIC values tend to be higher for a negative
    gram.

Which of your visuals above (1 through 5) is **most effective** at
helping to answer this question?

-   Visual 2.

Why?

-   The lollipop plots of the three different antibiotics type are
    stacked over one another, which makes it very easy to compare and
    distinguish among their MIC values.

#### Guiding Question 2

In 1974 *Diplococcus pneumoniae* was renamed *Streptococcus pneumoniae*,
and in 1984 *Streptococcus fecalis* was renamed *Enterococcus fecalis*
[2].

> Why was *Diplococcus pneumoniae* was renamed *Streptococcus
> pneumoniae*?

*Observations*

What is your response to the question above?

-   We can observe from Visual 4 that *Diplococcus pneumoniae* and
    *Streptococcus viridans* have similar Streptomycin MICs response,
    which further indicates that they might have similar biological
    structures. It's plausible that the naming of the antibiotics are
    closely related to their biological structures.

Which of your visuals above (1 through 5) is **most effective** at
helping to answer this question?

-   Visual 4.

Why?

-   According to Visual 4, *Diplococcus pneumoniae* is located very
    close to *Streptococcus viridans* and *Streptococcus hemolycitus*,
    which means their MIC values of streptomycin and penicillin are both
    pretty similar.

# References

<!-- -------------------------------------------------- -->

[1] Neomycin in skin infections: A new topical antibiotic with wide
antibacterial range and rarely sensitizing. Scope. 1951;3(5):4-7.

[2] Wainer and Lysen, "That's Funny..." *American Scientist* (2009)
[link](https://www.americanscientist.org/article/thats-funny)
